Uncovering underlying high-level musical content in the temporal domain \\ \small Leveraging self-supervised deep neural networks, inductive bias, and aural music skills to learn music embeddings with applications to MIR downstream tasks

### 2023-05-13 abstract/abstract.tex
Updates from Overleaf

The abstract should have at least 200 but not more than 600 words. Placed on the right page next to a blank left page.

\bigskip
Keywords: Music; Music Theory; Music Information Retrieval; Music Structure Analysis; Machine Learning; Neural Networks; Deep Learning

>>>>

Music, an integral part of human culture, offers a diverse yet complex field of study due to its intricate styles, mutable ground truth, and subjective nature. This research, conducted in collaboration with Epidemic Sound AB and the Music Technology Group at Universitat Pompeu Fabra, employs self-supervised learning and contrastive learning of musical representation to uncover the fundamental tonal structure of Western music.

Embracing a subtle novel approach, this process is facilitated through the application of Siamese networks with a triplet loss, mimicking human aural skills in interpreting music to discern musical elements, irrespective of the sonic qualities. The study replaces traditional acoustical features with deep audio embeddings, to compute high-level, sound-agnostic, and content-sensitive music identification.

Results XXXXXX

We strongly believe that this approach to learning music-informed embeddings can benefit almost every MIR downstream task. The potential implications of this research include advancements in intelligent music recommendation systems and efficient intellectual property rights enforcement, ultimately promoting innovation across the music industry.

\bigskip
Keywords: Music; Music Theory; Music Information Retrieval; Music Structure Analysis; Machine Learning; Deep Learning; Self-Supervised Learning;

### 2023-05-13 discussion/discussion.tex
Updates from Overleaf

@@

>>>>

XXXXXX

@@

In this work, we introduced a XXXX that leverages deep audio embeddings learned via self-supervised methods. Building on top of an existing algorithm based on XXX and framed within MSAF, we replaced traditional features with deep embeddings trained via XXXX for computing high-level music content identification.

### 2023-05-13 evaluation/subsections/beyond dsp.tex
Updates from Overleaf

>>>>

\input{figures/neural networks/r128 sphere}

\section{SALAMI dataset}
\subsection{Overview}

SALAMI (Structural Analysis of Large Amounts of Music Information) \cite{Smith2011DESIGNANNOTATIONS} aims to conduct massive structural analyses of different types of music. The project covers diverse music, from Western pop to Indian classical and from live to studio recordings.

\subsubsection{Formal Analysis Definition}
Formal analysis, in this context, doesn't refer to classifying the piece into a formal type (e.g., sonata, song, or canon) or reducing it into its Ursatz as in Schenkerian analysis. Instead, it means organizing and dividing the piece into specific sections and understanding how they relate.

%%%%%%%%%%

\subsubsection{Approaches to Music Analysis}

The method proposed combines aspects of the following three approaches, separating the organization of instrumentation, musical material, and formal function. This allows for the analysis of a wide variety of music pieces in a consistent way using a constrained vocabulary. It's important to note that music structure is often hierarchical, and the appropriate timescale for analysis can be hard to determine. The proposed method includes a few markers to partly address this issue.

Perceptual Approach: This involves listening for prominent harmonic or rhythmic boundaries in the piece to segment and apply labels to sections based on their similarity. However, this approach might not reflect the different structural roles of similar-sounding sections.

Functional Approach: This involves dividing the song into its functional parts, like verses, choruses, intros, outros, etc. This might result in the same music being labeled differently based on its function or different musical ideas having the same functional label.

Transcription Approach: This involves recording musical parameters like chord patterns and instrumentation directly. While this method may be less subjective, it may not provide a comprehensive view of the piece's structure.

### 2023-05-13 introduction/Subsections/Assumptions.tex
Updates from Overleaf

This thesis posits that high-level musical concepts manifest over time and remain invariant to specific attributes such as tonal quality, tempo, key, orchestration, signal noise, etc. This is akin to a sheet music composition performed in myriad ways using diverse instruments and styles while preserving its essence.

Abstract features, such as melody, harmony, rhythm, tempo, form, and expression, can be identified irrespective of waveform or production style. This approach offers an objective and comprehensive insight into a piece's musical content and meaning, analogous to how analyzing sheet music unveils a composition's underlying structure and intent.

Although a waveform might have limited sheet music representations that accurately encapsulate its musical components, a single sheet music composition can be performed infinitely using various instruments, voices, tempos, and interpretations. The distinct tonal quality of a waveform is significantly influenced by the specific instrument timbre or production technology employed in its creation, complicating the transcription of its sonic properties into precise sheet music notation.

Moreover, this distinct tonal quality is represented by high-level natural language concepts embedded within a rich cultural and sociological tradition.

\begin{figure}
    \centering
    \scalebox{0.9}{
    \includegraphics[width=\textwidth]{figures/images/Mahler 9 Giant Steps score.png}
        }
    \caption[Mahler's 9th Symphony, 2nd movement]{\small{A small excerpt from Mahler's 9th Symphony, 2nd movement: The melodic and harmonic contour propels through non-diatonic major thirds.}}
    \label{fig:mahler}
\end{figure}

\begin{figure}
    \centering
    \scalebox{0.9}{
    \includegraphics[width=\textwidth]{figures/images/giant steps score.png}
        }
    \caption[Giant Steps]{\small{John Coltrane's Giant Steps head: A testament to harmonic exploration, featuring rapid chord changes in non-diatonic major thirds.}}
    \label{fig:giant_steps}
\end{figure}

@@

\subsection{Beyond Digital Signal Processing (DSP)}

We argue that MIR research needs to incorporate a more balanced approach that considers the interdisciplinary nature of music and the importance of other domains beyond DSP.

>>>>

This thesis posits the existence of high-level musical concepts invariant to specific attributes such as tonal quality, tempo, key, orchestration, and signal noise that evolve and unfolds through time. This idea parallels the nature of sheet music, which maintains its essence despite being interpreted through diverse performances using various instruments and styles.

Abstract high-level features like melody, harmony, rhythm, tempo, form, and expression can be discerned regardless of waveform or production style. This approach provides an objective, comprehensive insight into a composition's musical content and meaning, much like analyzing sheet music uncovers its structural and intentional foundation.

It's worth noting that while a single waveform might have a limited number of sheet music representations that adequately capture its musical components, a single piece of sheet music can be interpreted in infinite ways through different instruments, voices, tempos, and interpretations. The distinct tonal quality of a waveform, heavily influenced by the specific timbre of the instrument or production technology, adds complexity to the transcription of its sonic properties into accurate sheet music notation.

Despite the differences in cultural contexts, stylistic elements, time signatures, key signatures, and sonic properties, we propose that figures \ref{fig:mahler} and \ref{fig:giant_steps} share a similarity. Both pieces carry the same musicological fingerprint, derived from the propulsion of their melodic and harmonic contours through non-diatonic major thirds.

This concept finds grounding in the theories of Kurt Koffka, a founder of the Gestalt school of psychology, who proposed a holistic approach to understanding complex forms, contrary to the structuralist practice of reducing mental processes into basic elements. His work, primarily embodied in the book, "Principles of Gestalt Psychology," established the foundation for Gestalt psychology and its applications to perception \cite{Koffka2013PrinciplesPsychology}.

When applied to music, the principles of Gestalt psychology offer insights into how we perceive, organize, and interpret musical patterns. They suggest that our minds perceive auditory input much like visual input, seeking patterns and structures. This understanding significantly contributes to the comprehension of cognitive processes in musical perception and organization, influencing music theory, cognition, and therapy \cite{Lerdahl1985AMusic}.

\begin{figure}[ht]
    \centering
    \scalebox{0.9}{
    \includegraphics[width=\textwidth]{figures/images/Mahler 9 Giant Steps score.png}
        }
    \caption[Mahler's 9th Symphony, 2nd movement]{\small{A small excerpt from Mahler's 9th Symphony, 2nd movement: The melodic and harmonic contour propels through non-diatonic major thirds.}}
    \label{fig:mahler}
\end{figure}

\begin{figure}[ht]
    \centering
    \scalebox{0.9}{
    \includegraphics[width=\textwidth]{figures/images/giant steps score.png}
        }
    \caption[Giant Steps]{\small{John Coltrane's Giant Steps head: A testament to harmonic exploration, featuring rapid chord changes in non-diatonic major thirds.}}
    \label{fig:giant_steps}
\end{figure}

@@

### 2023-05-13 methods/subsections/model architecture and training strategy.tex
Updates from Overleaf

\subsection{Problem to solve}

\subsection{Self-Supervised Learning (SSL)}

@@

This approach to learning is inspired by the way humans and animals learn from their environment, where a great deal of knowledge is acquired through observation and interaction without explicit instruction. SSL algorithms often involve tasks where the model learns to predict or reconstruct parts of the input data, such as predicting the next word in a sentence or completing an image with missing pixels.

@@

@@

Unknown invariance: Unknown invariances are transformations or changes the model has not been explicitly designed to handle. However, it still manages to be robust due to its learning capabilities. These invariances can emerge from the data and may not be known a priori. For example, a deep learning model might automatically learn to be invariant to certain lighting conditions or other factors not explicitly specified in its design.
\end{definition}

@@

\subsection{Contrastive Learning of Musical Representations (CLMR)}

@@

\subsection{Triplet Siamese Networks}

@@

$y_{c_o}(n)$ is the output for the $c_o$-th output channel at position $n$
$C_{\text{in}}$ is the number of input channels
$x_{c_i}(n)$ is the input for the $c_i$-th input channel at position $n$
$w_{c_o, c_i}(k)$ is the kernel weight for the $c_o$-th output channel and $c_i$-th input channel at position $k$
$K$ is the kernel size
The stride is set to 1, and padding is set to 1 (equal padding $\lfloor K/2 \rfloor$ on both sides)
s
    \item Batch normalization. We process a batch of data containing anchor, positive, and negative samples. We pad the waveforms to the same  maximum length among the collection, perform online triplet mining to find the hardest negatives and return the padded anchors, positives, and hardest negatives.

@@

We have adapted the original architecture for our experiments to our purpose, which consists of the final average pooling step.

@@

\begin{itemize}
  \item \(y_i\) represents the output tensor after the average pooling operation,
  \item \(x_{ij}\) represents the elements in the 2D input tensor,
  \item \(N\) is the number of elements along the second dimension of the input tensor,
  \item \(i\) iterates over the first dimension of the tensors, and
  \item \(j\) iterates over the second dimension of the input tensor.
\end{itemize}

@@

\subsubsection{Positive sample generation}
Given an anchor audio signal $A[n]$, we generate a positive signal $P[n]$ by applying a series of audio effects and adding noise. Let $g \in [-12, 0]$ represent the gain, $\Delta p \in [-1200, 1200]$ the pitch shift, $h_R[n]$ the impulse response of the reverb with parameters in $[0, 100]$, $h_C[n]$ the impulse response of the chorus with parameters determined by the specified ranges, $d \in [0, 30]$ the overdrive parameter, $\alpha \in [0.9, 1.1]$ the speed change factor, $\beta \in [0.9, 1.1]$ the stretch factor, $f_m \in [0.1, 100]$ the modulation frequency of the tremolo, and $d_m \in [1, 101]$ the depth of the tremolo.

>>>>

\subsubsection{Self-Supervised Learning (SSL)}

@@

This approach to learning is inspired by the way humans and animals learn from their environment, where a great deal of knowledge is acquired through observation and interaction without explicit instruction.

@@

Further understanding of the cognitive processes in music can be obtained through Jean Piaget's theory of cognitive development. Piaget suggests that children acquire knowledge through sensory and motor experiences in their early stages, with reasoning gradually evolving toward abstract ideas and deductive logic through the development of schemas—cognitive structures that underlie human knowledge and skill \cite{Huitt2003PiagetsDevelopment}. These schemas evolve through an equilibration mechanism that incorporates new information with pre-existing knowledge, involving processes of assimilation and accommodation \cite{audioselfsupsurvey}.

Although there are resemblances between assimilation and accommodation in Piaget's theory and backpropagation in Deep Learning (DL), these processes occur in distinct mechanisms and contexts. Assimilation and accommodation modify mental structures to incorporate new information, while backpropagation adjusts neural network weights to enhance output predictions.

@@

Unknown invariance: Unknown invariances are transformations the model has not been explicitly designed to handle. However, it still manages to be robust due to its learning capabilities. These invariances can emerge from the data and may not be known a priori. For example, a deep learning model might automatically learn to be invariant to certain lighting conditions or other factors not explicitly specified in its design.
\end{definition}

@@

\subsubsection{Contrastive Learning of Musical Representations (CLMR)}

@@

\subsubsection{Triplet Siamese Networks}

@@

$y_{c_o}(n)$ is the output for the $c_o$-th output channel at position $n$, $C_{\text{in}}$ is the number of input channels, $x_{c_i}(n)$ is the input for the $c_i$-th input channel at position $n$, $w_{c_o, c_i}(k)$ is the kernel weight for the $c_o$-th output channel and $c_i$-th input, channel at position $k$, and $K$ is the kernel size. The stride is set to 1 and padding to 1 (equal padding $\lfloor K/2 \rfloor$ on both sides).

\item Batch normalization. We process a batch of data containing anchor, positive, and negative samples. We pad the waveforms to the same  maximum length among the collection, perform online triplet mining to find the hardest negatives and return the padded anchors, positives, and hardest negatives.

We have intentionally chosen to use zero padding instead of array truncating in our experiment. This decision was based on our belief that preserving the integrity of the music data was more critical, despite the potential trade-off with computational efficiency during the normalization process.

@@

We have adapted the original architecture to our purpose, which consists of the final average pooling step.

@@

$y_i$ represents the output tensor after the average pooling operation, $x_{ij}$ represents the elements in the 2D input tensor, $N$ is the number of elements along the second dimension of the input tensor, $i$ iterates over the first dimension of the tensors, and $j$ iterates over the second dimension of the input tensor.

@@

\subsubsection{Positive sample generation}
Given an anchor audio signal $A[n]$, we generate a positive signal $P[n]$ by applying a series of amplitude, time-domain, frequency-domain, modulation, reverberation, and nonlinear effects with additive noise on top.

Let $g \in [-12, 0]$ represent the gain, $\Delta p \in [-1200, 1200]$ the pitch shift, $h_R[n]$ the impulse response of the reverb with parameters in $[0, 100]$, $h_C[n]$ the impulse response of the chorus with parameters determined by the specified ranges, $d \in [0, 30]$ the overdrive parameter, $\alpha \in [0.9, 1.1]$ the speed change factor, $\beta \in [0.9, 1.1]$ the stretch factor, $f_m \in [0.1, 100]$ the modulation frequency of the tremolo, and $d_m \in [1, 101]$ the depth of the tremolo.
